---
title: "class 5 Data Viz with ggplot 2"
author: "Kaif (A17497464)"
format: pdf
toc: true
---

## Background

There are many graphic systems in R for making plots and figures. These include so-called *"base R graphics* like the 'plot()' function and add on packages like **ggplot2**.

lets compare how we make a simple figure with these two systems:

We can use the in-built 'cars' dataset:

```{r}
head( cars )
```

```{r}
plot(cars)
```

Befor I can use ggplot2 I need to install it on my computer.
To do this we can use the fucntion 'install.packages("ggplot2")'

> **N.B.** We never run 'install.packages()' in our quarto doc (we run it once only in our R console) as it would re-install the package every time we render it
The main function in the **ggplot2** package is called 'ggplot()'

Once installed we need to load up the pakcage into our R brain: 

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers:

- the **data** (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how we want the plot drawn, e.g. points, lines, etc.)


```{r}
ggplot(cars) +
  aes(x=speed , y=dist)+
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance.

```{r}
ggplot(cars) +
  aes(x=speed , y=dist)+
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(title= "Stopping Dist Of Old Cars" ,
       x= "Speed (MPH)" , 
       y= "Distance (ft)")
```

```{r}
ggplot(cars) +
  aes(x=speed , y=dist)+
  geom_point() +
  geom_smooth(method= "lm", se= FALSE) +
  theme_bw() +
  labs(title= "Stopping Dist Of Old Cars" ,
       x= "Speed (MPH)" , 
       y= "Distance (ft)")
```


Q. Can you make the 'geom_smooth()' fucntion produce a linear straight line fit to the data and turn off the "gray" error region. 


## Gene expression figure 

Import the data to plot

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


```{r}
sum(genes$State == "up")
```

A useful new function in this context is the 'table()' fucntion:

```{r}
table(genes$State)
```

My first plot attempt

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col = State) +
  geom_point() +
  scale_color_manual(values = c("purple", "gray", "orange")) +
  theme_bw() +
  labs(x = "No drug", y = "Drug",
       title = "Expression changes upon GLP-1 inhibitor treatment")
```


## Going Further

Here we read the famous gapminder dataset

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

Q. How many entries (i.e. rows) are in this dataset

```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this dataset?

```{r}
length(table(gapminder$country) )
```

```{r}
length (unique(gapminder$country))
```

Lets's make our first plot of the entire dataset:

Plot of "gdpPercap" vs "lifeExp" colored by "continent"

```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point(alpha = 0.3)
```

I can add more layers to 'p'
```{r}
p +
  facet_wrap(~continent)
```

Make a plot for 1977 and 2007 only (notall the years in the dataset)


> Q. First use the **dplyr** package and the 'filter()' funtion from that package to extract the rows from the year 2007.



```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

```{r}
gap2007 <- filter(gapminder, year == 2007)
```

```{r}
gap_77_07 <- filter(gapminder, year %in% c(1977, 2007))
```

```{r}
table(gap_77_07$year)
```

```{r}
ggplot(gap_77_07) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point(alpha = 0.3) +
  facet_wrap(~year)
```
> Q. Make a histogram of lifeExp colored by continent (try using fill=continent or "col=continent)
> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder)+
  aes(lifeExp) +
  geom_histogram()
```

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill = continent) +
  geom_histogram()
```
```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill = continent) +
  geom_histogram(alpha = 0.6, position = "identity")
```

