---
title: "Class 6 : R Functions"
author: "Kaif Yenepoya PID (A17497464)"
format: pdf
toc: true
---

## Background

All functions in R have at least 3 things:

-   a *name* (we pick and use it to call the function)
-   input *arguments (one or more comma seperated inputs that go inside
    the brackets when we call the function), -the* body\* (the lines of
    R code that do work of the function)

## A first function

Here we will create a function to add some numbers. Let's call it
'add()'

Input arguments can either be "required" or "optional". The later have
fall-back default values that will be used if the useer does not specify
them.

> **Q1a**. Your first version of the function should add two input
> numbers together. For example, add(4, 7) should return 11. \[1 pt\]

```{r}
add <- function (x, y=0){
  x + y
}
```

Can we use our new function:

```{r}
add (10, 100)
```

```{r}
add(10)
```

> **Q1b** : For you second version, adapt your first function so it can
> take a single input vector or two inputs as before. For example,
> add(4, 7) and add( c(4,7,10) ). \[1 pt

```{r}
add <- function(x, y = NULL) {
  # If y is not given, assume x is a vector and sum all values
  if (is.null(y)) {
    return(sum(x))
  }
  
  # If y is given, add x and y
  return(x + y)
}
```

```{r}
add(4, 7)
add(c(4, 7, 10))
```

> **Q1c**: Finally, on your own (outside of class) create a third
> version of your function that can add any number of inputs that the
> user provides. For example, add(1, 2, 3, -4) should return 2. Hint:
> explore the ... (dots) argument or the base R function sum(). \[2
> pts\]

```{r}
add <- function(...) {
  # Collect all inputs into a list
  inputs <- list(...)
  
  # Unlist converts all inputs into one numeric vector, then sum adds them
  return(sum(unlist(inputs)))
}
```

```{r}
add(4, 7)
add(c(4, 7, 10))
add(1, 2, 3, -4)
```

We can explicitly set a **return** value output from a functin (rather
than the default last line) by using the 'return()' function call.

## A generate_dna() function

A useful function here is the "base R" 'sample()' fucntion

```{r}
sample(x=c("A", "C", "G", "T"), size=10, replace =TRUE)
```

We can use this to make a random nucleotide sequence iof we draw from
"A", "C", "G" and "T" ...

> **Q2a**: Write a function generate_dna() that returns a random DNA
> sequence of a length specified by the user. a. Your first version
> should return a multi-element vector of single character nucleotides.
> For example generate_dna(6) might return "A", "T", "T", "G", "A", "C".
> \[1 pt\]

```{r}
generate_dna <- function(len=10) {
  sample(x=c("A", "C", "G", "T"), size=len, replace =TRUE)
}
```

```{r}
generate_dna()
generate_dna(len=100)
```

> **Q2b** : b. Your second version should optionally be able to return
> either a multi-element vector of single character nucleotides (as
> before) or a single character string (not a vector of single letters
> but a singe vector of multiple letters). For example "AAGGCTG". \[1
> pt\]

Funtions that could be useful here are 'paste()', 'if()', 'cat()' and
'return()'

```{r}
generate_dna <- function(len = 10, as_string = FALSE) {
  dna <- sample(x = c("A", "C", "G", "T"), size = len, replace = TRUE)
  
  if (as_string) {
    return(paste(dna, collapse = ""))
  }
  
  return(dna)
}
```

```{r}
generate_dna(10)
generate_dna(10, as_string = TRUE)
```

> **Q2C** : c. Finally, create a final version of your function that
> prints out a FASTA format sequence with an id line indicating the
> sequence length.

```{r}
generate_dna <- function(len = 10) {
  # Generate a random vector of DNA nucleotides
  dna <- sample(x = c("A", "C", "G", "T"), size = len, replace = TRUE)
  
  # Collapse the vector into one single DNA string
  dna_string <- paste(dna, collapse = "")
  
  # Print the result in FASTA format with a sequence length ID
  cat(">len", len, "\n", dna_string, "\n", sep = "")
}
```

```{r}
generate_dna(9)
```

> **Q3**. Write a function generate_protein() that returns a random
> peptide/protein sequence of a length specified by the user. For
> example generate_protein(6) might return "WQRTAG".

```{r}
generate_protein <- function(len = 6) {
  # Store the 20 standard amino acid one-letter codes
  aa <- c("A", "R", "N", "D", "C", "E", "Q", "G", "H", "I",
          "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
  
  # Randomly sample amino acids to make a protein sequence
  protein <- sample(x = aa, size = len, replace = TRUE)
  
  # Collapse the amino acid vector into one single string
  protein_string <- paste(protein, collapse = "")
  
  # Return the final protein sequence
  return(protein_string)
}
```

```{r}
generate_protein(6)
```

## Generate random protein sequences of length 6 to 13

> **Q4**. Adapt and use your generate_protein() function to generate a
> series of random protein sequences ranging from 6 to 13 amino acids in
> length (one sequence per length). Take advantage of the base R
> function for() or sapply() so that you do not have to call
> generate_protein() eight times by hand.

```{r}
# Generate random protein sequences of length 6 to 13 in FASTA format
for (i in 6:13) {
  cat(">id.", i, "\n", generate_protein(i), "\n", sep = "")
}
```

## ARE OUR OPEPTIDES "UNIQUE IN NATURE" ?
> **Q5** : Take your FASTA-formatted peptides from Q4 and run them as a
> single BLASTp search against the Non-redundant protein sequences (nr)
> database at https://blast.ncbi.nlm.nih.gov/. For this question do not
> restrict the organism (leave the Organism field blank so that the full
> nr database is searched).

| length | Ide   | Cov | Unique |
|--------|-------|-----|--------|
| 6      | 100   | 100 | N      |
| 7      | 100   | 100 | N      |
| 8      | 100   | 100 | N      |
| 9      | 100   | 89  | Y      |
| 10     | 100   | 90  | Y      |
| 11     | 81.82 | 100 | Y      |
| 12     | 88.89 | 75  | Y      |
| 13     | 77.78 | 69  | Y      |
|        |       |     |        |

> **Q5a** : At which sequence length do your randomly generated peptides start to look “unique in nature”
(i.e. no 100% coverage + 100% identity hit)? [1 pt]

The randomly generated peptides started to look unique in nature at length 9, because this was the first peptide without a hit showing both 100% query coverage and 100% identity.

> **Q5b** : Speculate why very short random peptides are almost always found in nr, while longer ones
typically are not. Your answer should refer both to the size of the sequence space (20𝐿 for a
peptide of length 𝐿) and to the size of the known protein universe. [1 pt

The probability that very short peptides will be detected within nr is higher due to the relatively limited size of the sequence space. The exponential increase in the size of the sequence space, 20^L, will soon become much larger than the known size of the protein universe, decreasing the possibility of exact matches.

## Connecting our findings to immunulogy

> **Q6** : In 3–6 sentences total and using your Q5 data and the reasoning from Q5b, what do you think
this minimum length is and why might it be a bad design choice for the immune system to present
very short peptides? [3 pt]

The lower limit of the length of peptides is estimated to be approximately nine amino acids since, as indicated by our Q5 results, peptides between six to eight amino acids retained 100% sequence match identity and coverage, whereas peptides above length nine became unique in their nature. The use of very short peptides for antigen recognition and presentation is a poor idea as there will be plenty of them in already known proteins, making them hard to discriminate between self and non-self.